Approximation as Differentiation: Notes

### 2025-05-26 main/category-of-families.tex
Initial import of FinVect, but too lengthy..have specialised to the reals, but should also specialise to the finite case too.

\todo{Maybe unpack why these are true}

\begin{proposition}
\label{prop:Grothendieck:fam-inherits}
\item
\begin{enumerate}
\item If $\cat{C}$ is locally small then so is $\Fam(\cat{C})$.
\item If $\cat{C}$ has binary products then so does $\Fam(\cat{C})$.
\end{enumerate}
\end{proposition}

@@

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. We recapitulate the basic idea here. For flat manifolds like
$\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically isomorphic to $\RR^n$, so
the forward and backward derivatives of a differentiable function $f: \RR^m \to \RR^n$ at a point $x$ have the
following form:

>>>>

\begin{proposition}
\label{prop:Grothendieck:fam-inherits}
\item \todo{Maybe unpack why these are true}
\begin{enumerate}
\item If $\cat{C}$ is locally small then so is $\Fam(\cat{C})$.
\item If $\cat{C}$ has binary products then so does $\Fam(\cat{C})$.
\end{enumerate}
\end{proposition}

@@

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. We recapitulate the basic ideas here, starting with the
category $\FinVect$ of finite vector spaces over $\RR$.

\subsubsection{Category of finite-dimensional vector spaces}
\label{sec:categories-with-biproducts:fdvect}

\begin{definition}[Vector space over $\RR$]
We define vector spaces over the reals, with addition of real numbers $a, b$ written $a + b$ and
multiplication written $a \mult b$. Then a \emph{vector space $V = (V, +, \mult)$ over $\RR$} (overloading $+$
and $\mult$) is a set $V$ equipped with addition of vectors $+: V^2 \to V$ and scalar multiplication $\mult:
\RR \times V \to V$ where $(V,+)$ is an abelian group (with identity element $0_V$ and additive inverse $-v$
for any $v \in V$) and the vector operations are compatible with the field operations in that the following
equations hold:

\vspace{-3mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{align*}
   1 \mult v &= v \\
   (a \mult b) \mult v &= a \mult (b \mult v)
\end{align*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{align*}
   (a + b) \mult v &= (a \mult v) + (b \mult v) \\
   a \mult (u + v) &= (a \mult u) + (b \mult v)
\end{align*}
\end{center}
\end{minipage}
\end{definition}

Up to isomorphism a finite-dimensional vector space over $\RR$ of dimension $n \in \Nat$ is the vector space
$(\RR^n, +, \mult)$ where $+$ and $\mult$ are defined component-wise.

\begin{definition}[Category $\FinVect$]
Define $\FinVect$ to be the category which has:
\begin{itemize}
\item as objects $V, W$ all pairs of natural numbers $n$ and $n$-dimensional vector spaces $(\RR^n, +, \mult)$
\item as morphisms $f: V \to W$ all functions $f$ satisfying $f(u + v) = f(u) + f(v)$ and $f(a \mult v) = a
\mult f(v)$.
\end{itemize}
\end{definition}

$\FinVect$ is enriched in $\Ab$, the category of abelian groups. For any $V, W$ we have that $\FinVect(V,W)$
is an abelian group, where:
\begin{enumerate}
\item $0_{V,W}(v) = 0_W$
\item $(f + g)(v) = f(v) + g(v)$
\item $(-f)(v) = -f(v)$
\end{enumerate}
\noindent with composition bilinear by (2) and addition of morphisms computed component-wise. The direct sum
$V \biprod W = (V \times W, +, \mult)$, with $+$ and $\mult$ defined component-wise, is a biproduct $(V
\biprod W, \biinj_V, \biinj_W, \pi_1, \pi_2)$ with $\biinj_V = \prodM{\id_V}{0_{V,W}}$ and $\biinj_W =
\prodM{0_{V,W}}{\id_W}$.

\subsubsection{Derivatives of Differentiable Functions}

For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
\RR^n$ at a point $x$ have the following form:
